Decline for some, resilience for others: changes in the prevalence of microbial symbionts over 150 years of climate change

### 2022-12-15 Manuscript/endo_herbarium.tex
updates to the intro

\title{Idiosyncratic responses of microbial mutualisms over 150 years of climate change}

@@

	% The journal does not have numbered sections in the main portion of
	% articles. Please refrain from using section references (à la
	% section~\ref{section:CountingOwlEggs}), and refer to sections by name
	% (e.g. section ``Counting Owl Eggs'').
Global climate change involves hetergenous changes in multiple climate drivers \citep{IPCC2012managing}. 
It is of great interest to predict species' responses to these drivers.
Novel conditions .Predicting how species' respond to these drivers, it is important to understand how and under what conditions interactions between species will alter expected responses. 
Incorporating interactions may lead to different predictions than those based on single species responses; for most species, we do not know whether their interactions will facilitate or hinder movement and adaptation to novel conditions \cite{gilman2010framework}. 
Natural history collections, which were originally collected to study and preserve taxonomic diversity, present an exciting opportunity to explore long-term changes in ecological interactions across broad spatial scales \citep{meineke2018unrealized}. 
These collections primarily comprise physical specimens of organisms collected along with information about the time and place of collection. 
These natural history collections, built and maintained by the contributions of thousands of scientists, are invaluable time machines. 
The specimens they contain are samples from dynamically changing environmental conditions through time and space.
These specimens can preserve physical legacies of ecological dynamics and species' interactions. 
Recent work using collections has documented changes in phenology \citep{willis2017old, park2019herbarium, berg2019examination}, as well as pollination \citep{pauw2011reconstruction, duan2019century} and changes in negative interactions such as herbivory \citep{meineke2019herbarium}, but to date, focus has been lacking on long-term changges in reciprocally positive interactions, i.e. mutualisms.

@@

Early research on \emph{Epichloë} fungal endophytes, symbionts of cool-season grasses, used herbarium specimens to describe the broad taxonomic diversity of hosts \citep{white1985endophyte}. 
These endophytes are vertically-transmitted, growing throughout the aboveground tissue and into the seeds of their hosts, and consequently are preserved along with the plant on herbarium sheets. 
Vertical transmission links the fitness of host and symbiont together meaning that, over time, endophytes that provide fitness benefits should rise in prevalence within the population \citep{fine1975vectors, douglas1998host, rudgers2009fungus}. 
This leads to the prediction that interaction prevalence should generally be high where there are strong fitness benefits, and that prevalence should track changes in environmental drivers that limit fitness. 
Environmental stress is likely particularly consequential at species' range edges wherer population growth rates are expected to be near replacement levels \cite{}. 
And so understanding the ways that endophytes may mediate stress is important for understanding the factors setting species' range limits. 
The interaction prevalence should respond to the fitness effects of the mutualism, and the prevalence of endophytes across the range determines their actual contribution to setting range limits. 
In addition, the rate of environmental change is variable across space across species ranges \citep{IPCC2012managing}, and local changes in endophyte prevalence likely depend on these rates.

And so, we are taking advantage of the spatial and temporal sampling within herbaria to ask how change in spatial and temporal environmental drivers alter the prevaence of a widespread mutualistic fungal endophyte symbiosis between Epichloë fungi and grasses. 
Specifically,  we ask (1) how variation in temperature and precipitation correlate with endophyte prevalence across space and over time, and (2) does greater rates of envirnmental change drive greater changes in endophyte prevalence.

>>>>

\title{Decline for some, resilience for others: changes in the prevalence of microbial symbionts over 150 years of climate change}

@@

	% The journal does not have numbered sections in the main portion of
	% articles. Please refrain from using section references (à la
	% section~\ref{section:CountingOwlEggs}), and refer to sections by name
	% (e.g. section ``Counting Owl Eggs'').
Global climate change involves hetergenous changes in multiple climate drivers \citep{IPCC2012managing}. 
Predicting species' responses to these drivers, whether species will move, adapt, or potentially go extinct is a major goal of basic ecological research and conservation management. 
Across the tree of life, species interact with macro- and microscopic symbionts which impact their growth and performance. 
These biotic interactions likely modulate responses to changing environmental conditions, yet for most species, we do not know whether they will facilitate or hinder movement and adaptation to novel conditions \cite{gilman2010framework}. Predictions incorporating biotic interactions are likely different than those based on single species responses, but empirical assements of these predictions have been limited by the lack of long-term and spatially extensive datasets of biotic interactions.

Natural history collections, which were originally collected to study and preserve taxonomic diversity, present an exciting opportunity to explore long-term changes in ecological interactions across broad spatial scales \citep{meineke2018unrealized}. 
These collections primarily comprise physical specimens of organisms collected along with information about the time and place of collection. 
These natural history collections, built and maintained by the contributions of thousands of scientists, are invaluable time machines. 
The specimens they contain are samples from dynamically changing environmental conditions through time and space.
These specimens can preserve physical legacies of ecological dynamics and species' interactions. 
Recent work using collections has documented changes in phenology \citep{willis2017old, park2019herbarium, berg2019examination}, as well as pollination \citep{pauw2011reconstruction, duan2019century} and changes in negative interactions such as herbivory \citep{meineke2019herbarium}, but to date, focus has been lacking on long-term changges in reciprocally positive interactions, i.e. mutualisms.

@@

Early research on \emph{Epichloë} fungal endophytes, symbionts of cool-season grasses, used herbarium specimens to describe the broad taxonomic diversity of hosts \citep{white1985endophyte}. 
These endophytes are vertically-transmitted, growing throughout the aboveground tissue and into the seeds of their hosts, and consequently are preserved along with the plant on herbarium sheets. 
Vertical transmission links the fitness of host and symbiont together meaning that, over time, endophytes that provide fitness benefits should rise in prevalence within the population \citep{fine1975vectors, douglas1998host, rudgers2009fungus}. 
This leads to the prediction that interaction prevalence should generally be high where there are strong fitness benefits, and that prevalence should track changes in environmental drivers that limit fitness. 
Environmental stress is likely particularly consequential at species' range edges where population growth rates are expected to be near replacement levels \cite{}. 
Understanding the ways that endophytes may mediate stress is important for understanding the factors setting species' range limits. 
The interaction prevalence should respond to the fitness effects of the mutualism, and the prevalence of endophytes across the range determines their actual contribution to setting range limits. 
In addition, the rate of environmental change is variable across space across species ranges \citep{IPCC2012managing}, and local changes in endophyte prevalence likely depend on these rates. 
We are taking advantage of the spatial and temporal sampling within herbaria to ask how the prevalence of a widespread mutualistic fungal endophyte symbiosis between Epichloë fungi and grasses has changed in response to changing climate drivers.
Specifically, the objectives of this study are to (i) quantify spatial and temporal change in endophyte prevalence, (ii) identify climate drivers that can explain changes in prevalence, and (iii) assess how the rate of environmental change influences changes in the prevalence of symbionts.
